---
title: "Modelling the probability of earthquakes (M >= 5.0) in North Anatolian Fault Zone"
date: 08/25/2022
author: Murat Koptur
execute:
  freeze: auto
  echo: false
  warning: false
  error: false
format: 
   html:
     df-print: paged
listing: true
---

```{r}
library(readxl)
library(sf)
library(raster)
library(dplyr)
library(spData)
library(tmap)    
library(leaflet)
library(ggplot2)
library(fitdistrplus)

set.seed(1234)
```

```{r}
data_with_coordinates <- read_excel("data.xlsx", 
    sheet = "earth", col_types = c("date", 
        "numeric", "numeric", "numeric", 
        "numeric", "skip"))
data_diff_between_eq <- read_excel("data.xlsx", 
    sheet = "diff")
data_count_by_year <- read_excel("data.xlsx", 
    sheet = "byyear")
```

## EDA

Let's look and visualize the historical earthquake data.

### Spatial data

```{r}
data_with_coordinates
```

```{r}
tur_shp <- st_read("gadm41_TUR.gpkg", layer = "ADM_ADM_0")
spdf <- SpatialPointsDataFrame(coords = data_with_coordinates[c(3,2)], data = data_with_coordinates)

tm_shape(tur_shp) + tm_fill() + tm_shape(spdf) + tm_dots(size=0.5, col = "magnitude", alpha=0.4)
```

### Time Span between Earthquake Occurrences

```{r}
summary(data_diff_between_eq)
```

```{r}
p <- ggplot(data_diff_between_eq, aes(x=diff)) + geom_histogram(aes(y = ..density..)) + geom_density()
p
```

```{r}
p2 <- ggplot(data_diff_between_eq, aes(x=diff)) + geom_boxplot()
p2
```

### Earthquake Count By Year

```{r}
p3 <- ggplot(data_count_by_year, aes(x=year, y=count)) + geom_line()
p3
```

## Modelling the probability

Let's fit Weibull distribution to distribution of days between two earthquakes occurred successively.

```{r}
plotdist(data_diff_between_eq$diff, demp = TRUE)
```

```{r}
descdist(data_diff_between_eq$diff, boot = 10000)
```

```{r}
# add all data points to 0.1 for avoiding zero division errors
data_diff_between_eq$diff <- data_diff_between_eq$diff + 0.01

#lam0 <- quantile(data_diff_between_eq$diff, p=0.632)

#Fh <- ppoints(sort(data_diff_between_eq$diff))
#k0 <- lm(log(-log(1-Fh))~log(data_diff_between_eq$diff))$coefficients[2]

wei.fit <- fitdist(data_diff_between_eq$diff, "weibull")
```

```{r}
print(wei.fit$convergence)
```

```{r}
summary(wei.fit)
```

```{r}
plot(wei.fit)
```

Let's calculate mean occurence period of earthquakes which have magnitudes equal or bigger than 5 (simulation and theoretical mean):

```{r}
shape.v <- as.numeric(wei.fit$estimate[1])
scale.v <- as.numeric(wei.fit$estimate[2])

simulated_data <- rweibull(100000, shape = shape.v, scale = scale.v)
print("Simulated mean:")
print(mean(simulated_data))
print("Theoretical mean:")
print(scale.v * gamma(1 + (1 / shape.v)))
```

It is expected to have another earthquake having magnitude equal to 5 or above are average `r round(scale.v * gamma(1 + (1 / shape.v)))` days later than the preceding one.

Let's plot the CDF:

```{r}
plot(ecdf(simulated_data), xlim=c(0, 6000))
```

If we look the data, the last earthquake was occurred at 2006-10-24, so 5753 days passed since last earthquake was occurred. The risk of an earthquake happening today is `r round(ecdf(simulated_data)(5753), digits=2)*100` %.

## References

-   Earthquake historical data was downloaded from: <http://deprem.afad.gov.tr/depremkatalogu>

-   Map shape file was downloaded from: <https://gadm.org/>

-   Weibull distribution in earthquake probability modelling: <https://www.researchgate.net/publication/26485414_Probabilistic_Prediction_of_the_Next_Earthquake_in_The_Nafz_North_Anatolian_Fault_Zone_Turkey_Kuzey_Anadolu_Fay_Zonunda_Nafz_Gelecek_Depremlerin_Olasiliksal_Tahmini>

-   Weibull distribution fitting: <https://stats.stackexchange.com/questions/230937/how-to-find-initial-values-for-weibull-mle-in-r>
