---
title: "Don't Try to Forecast Everything: Predictability of Time Series"
date: 09/01/2022
author: Murat Koptur
draft: true
execute:
  freeze: auto
  echo: false
  warning: false
  error: false
format: 
   html:
     df-print: paged
listing: true
---

Most of time series analyses start with investigating series, autocorrelation and 
partial autocorrelation plots. Then one estimates different time series models
(like ARIMA, GARCH, State-space models) and performs model checks.

But no one asks whether that series is predictable or not.

We'll look at a few handy tools that give more information about our time series.

```{r}
library(DChaos)
library(TSstudio)
```


## Data

We'll use some example time series:

* Monthly Airline Passenger Numbers 1949-1960 (AirPassengers)

  ```{r}
  data(AirPassengers)
  ts_plot(AirPassengers)
  ```
  
* Level of Lake Huron 1875-1972 (LakeHuron)

  ```{r}
  data(LakeHuron)
  ts_plot(LakeHuron)
  ```
  
* Simulated time-series data from the Logistic map with chaos

  ```{r}
  LogisticTs <- logistic.sim(a=4, s=0.5, n=100)
  ts_plot(LogisticTs)
  ```
Let's look the tools.

## Lyapunov Exponent



## References

* DChaos, https://cran.r-project.org/web/packages/DChaos/index.html
* statcomp, https://cran.r-project.org/web/packages/statcomp/index.html
* pracma, https://cran.r-project.org/web/packages/pracma/index.html
* tseriesChaos, https://cran.r-project.org/web/packages/tseriesChaos/index.html
* Daniel F. McCaffrey , Stephen Ellner , A. Ronald Gallant & Douglas W. Nychka (1992) Estimating the
Lyapunov Exponent of a Chaotic System with Nonparametric Regression, Journal of the American Statistical Association,
87:419, 682-695

